---
title: "Economic and Human Loss by Storms in the U.S."
author: "[John H. Poplett](mailto:john.poplett@acm.org)"
date: "`r Sys.Date()`"
output: html_document
---
# Synopsis
This study shows some of the more dire consequences of storms in the US, including economic loss and loss of life. It further illustrates the kinds of storms that have the greatest impact on human lives, economically and health wise. The data is drawn from the [storm events database](http://www.ncdc.noaa.gov/stormevents/) available through [Climate Data Online](http://www.ncdc.noaa.gov/cdo-web/), a national resource from the [National Centers for Environmental Information](http://www.ncdc.noaa.gov) (NCEI).

# Data Processing
To make the data pristine and produce meaningful results, steps were taken to "clean" the raw data, especially to eliminate data that has been mislabeled. The code sections below show how the data is loaded. The code converts the event type column, "EVTYPE", to upper-case, corrects the event type column, "EVTYPE", for whitespace errors and then weeds out the proportionally small amount of EVTYPE values that don't conform to the canonical naming conventions described by the NCEI for event types.

## Define Constants
Define constants to get the ball rolling.

```{r constants, echo = TRUE, cache = TRUE, message = FALSE}
storm.event.data.URL <- "https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2FStormData.csv.bz2"
storm.event.data.file.name <- "repdata-data-StormData.csv"
```

## Data Loading and EVTYPE Value Clean-up
The code performs the following steps to process the data; it:

* downloads compressed Compressed data from this Internet [location](`r storm.event.data.URL`).
* converts EVTYPE values to upper-case
* reduces multiple consecutive whitespaces in EVTYPE values to a single space
* trims EVTYPE values of Leading and trailing whitespace

```{r load, echo = TRUE, cache = TRUE}
storm.data <- read.csv(storm.event.data.file.name)
storm.data$EVTYPE <- as.character(storm.data$EVTYPE)
storm.data$EVTYPE <- toupper(storm.data$EVTYPE)
storm.data$EVTYPE <- gsub("\\s+", " ", storm.data$EVTYPE, perl = TRUE)
storm.data$EVTYPE <- sub("^\\s+", "", storm.data$EVTYPE, perl = TRUE)
storm.data$EVTYPE <- sub("\\s+$", "", storm.data$EVTYPE, perl = TRUE)
```

## Event Type Filtering

```{r evtype-filtering}
library(dplyr)
evtypes <- data.frame(evtype = c("ASTRONOMICAL LOW TIDE", "AVALANCHE",
  "BLIZZARD", "COASTAL FLOOD", "COLD/WIND CHILL", "DEBRIS FLOW",
  "DENSE FOG", "DENSE SMOKE", "DROUGHT", "DUST DEVIL", "DUST STORM",
  "EXCESSIVE HEAT", "EXTREME COLD/WIND CHILL", "FLASH FLOOD", "FLOOD",
  "FROST/FREEZE", "FUNNEL CLOUD", "FREEZING FOG", "HAIL", "HEAT",
  "HEAVY RAIN", "HEAVY SNOW", "HIGH SURF", "HIGH WIND", "HURRICANE (TYPHOON)",
  "ICE STORM", "LAKE-EFFECT SNOW", "LAKESHORE FLOOD", "LIGHTNING",
  "MARINE HAIL", "MARINE HIGH WIND", "MARINE STRONG WIND", "MARINE THUNDERSTORM WIND",
  "RIP CURRENT", "SEICHE", "SLEET", "STORM SURGE/TIDE", "STRONG WIND",
  "THUNDERSTORM WIND", "TORNADO", "TROPICAL DEPRESSION", "TROPICAL STORM",
  "TSUNAMI", "VOLCANIC ASH", "WATERSPOUT", "WILDFIRE", "WINTER STORM",
  "WINTER WEATHER"))

evtype.values.before <- length(unique(storm.data$EVTYPE))
storm.data.rows.before <- nrow(storm.data)

storm.data <- storm.data %>% filter(storm.data$EVTYPE %in% evtypes$evtype)

evtype.values.after <- length(unique(storm.data$EVTYPE))
storm.data.rows.after <- nrow(storm.data)
```

Before filtering, there were `r evtype.values.before` evtype values distributed over `r prettyNum(storm.data.rows.before, big.mark = ",")` rows. After filtering, there are `r evtype.values.after` evtype values distributed over `r prettyNum(storm.data.rows.after, big.mark = ",")` rows.

# Results
Results are divided into two parts, concerning respectively:

* human harm
* economic harm

## Events Most Harmful to Population Health
The following code identifies and displays the events that are most harmful to humans in terms of the total count of fatalities, injuries and incidents (the sum of fatalities and injuries) that are attributed to individual storm events.

```{r most-harmful, fig.height = 8, fig.width = 16}
library(dplyr)
library(reshape2)
library(ggplot2)

storm.data.summary <- storm.data %>%
  group_by(EVTYPE) %>%
  summarise_each(funs(sum), FATALITIES, INJURIES) %>%
  arrange(desc(FATALITIES))

storm.data.summary$INCIDENTS <- storm.data.summary$FATALITIES + storm.data.summary$INJURIES

fatalities <- head(storm.data.summary, n = 5)
injuries <- head(storm.data.summary %>% arrange(desc(INJURIES)), n = 5)
incidents <- head(storm.data.summary %>% arrange(desc(INCIDENTS)), n = 5)

fatalities <- fatalities %>% melt(id.vars = "EVTYPE", measure.vars = "FATALITIES")
injuries <- injuries %>% melt(id.vars = "EVTYPE", measure.vars = "INJURIES")
incidents <- incidents %>% melt(id.vars = "EVTYPE", measure.vars = "INCIDENTS")

most.harmful <- rbind(fatalities, injuries, incidents)

ggplot(data=most.harmful, aes(x=EVTYPE, y=value, fill=variable)) + 
  geom_bar(colour="black", stat="identity",
           position=position_dodge(),
           size=.3) +                        # Thinner lines
  scale_fill_hue(name="Metric") +      # Set legend title
  xlab("Storm Event") + ylab("Count") + # Set axis labels
  ggtitle("Human Harm") +     # Set title
  theme_bw()

```

# Notes for Data Scientists

## Data Cleaning
Other approaches to data cleaning were considered including efforts to correct various kinds of mislabeling errors, e.g.:

* typos (e.g. "THUNERSTORMS")
* word swaps (e.g. "high" for "heavy")
* synonyms (e.g. "precipitation" for "rain" or "driest month" for "drought")

## Adjusting for Inflation
It is possible to consider adjusting historic damage amounts for inflation to normalize them. This approach was considered but not taken since, for the most part, the relative value of currency in a given year will be the same for all storm event types. So despite inflation, the proportions should remain roughly on par.
